---
title: "Data handling"
author: "Lisa Hülsmann"
date: "18 Mai 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading excel files

Let's import a new dataset from excel. This is possible with the package readxl. 
```{r}

library(readxl)
test = read_excel("../data/test.xlsx")
```

R Markdown files use their own working directory. So you have to provide the path relative to the location of the RMarkdown file!!!

```{r}
str(test) # this is a tibble and a data.frame

```

Data imported with read_excel() is stored as a tibble and a data.frame. For typical analysis we have to convert it to data.frame.

```{r}
testDF = as.data.frame(test)
str(testDF)
```

# Solution of the exercise

## Read data into R

First, you read in the files using read.csv(). You have to modify the separator. 

```{r}
treedata = read.csv("../data/treedata.csv", sep = ";")
species = read.csv("../data/species.csv", sep = ";")
# View(species)
```

Note that the View() function is not printed in the Rmarkdown output.

## Dataset properties

To get an overview of the structure of a dataset we use str().
```{r}
str(treedata)
```

The dataset has four colums: species code, dbh (diameter at breast height [cm]), heigth [m] and remarks. We would expect the following classes:

species code: integer
dbh: numeric
height: numeric
rem: factor

But dbh is not numeric but a factor. 

## Change to numeric and remove weird values

Searching for the X
```{r}
select = treedata$dbh == "X"

# alternatives
select = treedata$dbh %in% LETTERS | treedata$dbh %in% letters

# or if even two letters are possible
select = grepl("^[A-Za-z]+$", treedata$dbh, perl = T)

```

replace it in the vector and convert to numeric.
Note that you have to convert to character before!!!
Otherwise the facto levels will be the numbers you obtain.

```{r}
treedata$dbh[select] = NA
treedata$dbh = as.numeric(as.character(treedata$dbh))
```

## How many NA values in height?

```{r}
sum(is.na(treedata$height)) # sum of a logical vector = number of TRUE
```
There are 221 NA values.



## Implausible values

```{r}
pairs(treedata[, c("dbh", "height")])
```

One height and one DBH are very strange: a tree larger that 100 m? or a dbh of 2000 cm?
Lets put these to NA.

```{r}
select = treedata$height > 60 & !is.na(treedata$height) # larger 60 and not NA
treedata$height[select] = NA

select = treedata$dbh > 1000 & !is.na(treedata$dbh) # larger 60 and not NA
treedata$dbh[select] = NA
pairs(treedata[, c("dbh", "height")])
```



## add information from a lookup table

Use match() or merge()
```{r}
treedata$specie_english = species$species_english[match(treedata$species_code, species$species_code)]
```



## Summarizing data

```{r}
# install.packages("dplyr")
library(dplyr)

summary = treedata %>% 
  group_by(specie_english) %>% 
  summarise(meanDBH = mean(dbh, na.rm = T), 
            meadianDBH = median(dbh, na.rm = T))


```



